# Results
```{r}
library(ggplot2)
library(dplyr)
library(forcats)

data <- read.csv("./data/data_clean.csv", stringsAsFactors = TRUE)
```
Explanation of variables:MARSTAT is the marriage status, LIVARAG is Living arrangements
SERVICE is Type of treatment service/setting, METHUSE is whether the use of opioid medications, DSMCRIT is the reason for a client's encounter or treatment
```{r}
head(data)
```
First we want to see the basic distribution for the age of data:
```{r}
ggplot(data, aes(x = AGE)) + 
  geom_bar(stat = "count", fill = "skyblue") + 
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  theme_minimal() +
  labs(x = "Age", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
From which we can see most of individuals are within 18-64 age.
And gender:
```{r}
ggplot(data, aes(x = GENDER)) + 
  geom_bar(stat = "count", fill = "#FF5733") + 
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  theme_minimal() +
  labs(x = "Age", y = "Count")
```
The variable SUB1 identifies the client's primary substance abuse at admission. The basic distribution is:
```{r}
data$SUB1 <- fct_infreq(data$SUB1)
ggplot(data, aes(x = SUB1)) + 
  geom_bar(stat = "count", fill = "skyblue") + 
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  theme_minimal() +
  labs(x = "Age", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


